Re-analysing the data from Moffatt et al. (\protect\hyperlink{ref-moffatt_inner_2020}{2020}): A textbook illustration of the absence of evidence fallacy

### 2020-09-30 manuscript.Rmd
minor updates in manuscript

---
title: "Re-analysing the data from @moffatt_inner_2020: A textbook illustration of the absence of evidence fallacy"
shorttitle: "The absence of evidence fallacy"
author: 
  - name: Ladislas Nalborczyk
    affiliation: 1
    corresponding: yes
    address: GIPSA-lab, CNRS, Univ. Grenoble Alpes, 11 Rue des Mathématiques, 38400 Saint-Martin-d'Hères, France
    email: ladislas.nalborczyk@gipsa-lab.fr
affiliation:
  - id: 1
    institution: Univ. Grenoble Alpes, CNRS, Grenoble INP, GIPSA-lab, 38000 Grenoble, France
abstract: "@moffatt_inner_2020 reported the results of an experiment (N = 26 in the final sample) comparing the facial (surface) electromyographic correlates of mental rumination and distraction, following an experimentally induced stressor. Based on the absence of significant difference in the perioral muscular activity between the rumination and distraction conditions, @moffatt_inner_2020 concluded that *self-reported* inner experience was unrelated to peripheral muscular activity as assessed using surface electromyography. We suggest this conclusion is hasty and based on waggly evidence. Indeed, concluding on the absence of an effect based on a low-powered non-significant p-value is strongly problematic/uninformative. Moreover, the relation between self-reports and physiological measures was not *directly* assessed, but only indirectly inferred from differences (or absence thereof) in group means. Given the ample inter-individual variability in these measures (as suggested by our reanalysis), we think inferring the individual-level relation between self-reports and physiological measures from group means is inappropriate. Given these limitations, we conclude that it is unclear whether the target article adds to the current/extent knowledge and we suggest ways forward, both from a theoretical and from a methodological perspective. Complete source code, reproducible analyses, and figures are available at https://github.com/lnalborczyk/inner_experience_EMG."
keywords: NHST, Bayesian, fallacy, reanalysis, inner speech, rumination, electromyography
csl: csl/apa7.csl
classoption: man
linenumbers: yes
floatsintext: yes
link-citations: true
numbersections: false
bibliography: [bib/library.bib, bib/r-references.bib]
nocite: | 
  @R-base, @R-knitr, @R-rmarkdown, @R-tidyverse, @R-papaja, @R-wordcountaddin,
  @R-here
output:
  papaja::apa6_pdf:
    keep_tex: true
    highlight: tango
    latex_engine: xelatex
editor_options: 
  chunk_output_type: console
---

```{r setup, include = FALSE, message = FALSE, warning = FALSE}
library(wordcountaddin)
library(BayesFactor)
library(ggbeeswarm)
library(rmarkdown)
library(tidyverse)
library(papaja)
library(readxl)
library(knitr)
library(brms)
library(here)
library(pwr)

@@

@@

<!-- NB: You can add comments using these tags -->

@@

The activity of silently talking to oneself or "inner speech" is a foundational ability... despite its multiple adaptive functions in everyday life, inner speech can go awry and leads to sustained negative... These inner speech "dysfunctions" [for reviews, see @alderson-day_inner_2015; @perrone-bertolotti_what_2014; @loevenbruck_cognitive_2018]...

Given the predominantly verbal nature of rumination [], we previously proposed to study rumination as other forms of inner speech have been studied in the past, namely using surface electromyography and motor interference protocols [e.g., @nalborczyk_orofacial_2017; @nalborczyk_understanding_2019; @nalborczyk_can_2020; @nalborczyk_articulatory_2020]...

We have previously shown that... However, it was unclear... therefore, the extension of our study by @moffatt_inner_2020, which consisted by including a distraction control group, is more than welcome... However...

The main conclusion from @moffatt_inner_2020 is that inner experience between induced rumination and distraction differs "without a change in electromyographic correlates of inner speech". In other words, their conclusion is that inner experience is unrelated (or loosely related) to the electromyographic correlates of inner speech, which are thought to be represented mostly by the EMG amplitude recorded over the OOI and OOS muscles. However, for this in-sample observation to be of interest in an out-of-sample context (i.e., to be informative of other non-observed individuals, or said otherwise, to bring information about the population), this absence of difference has to be based on sufficiently powered sample size (given the target effect size) as well as on reliable measures. Moreover, a simple visual exploration of the data reveals important variability between individuals in the main effect of interest. That is, some participants had higher perioral (OOS and OOI) muscular activity in the rumination condition than in the distraction condition, and some other participants showed the reverse pattern. This suggests unexplored variation in the determinants of this effects (e.g., the content of the inner experience). Indeed, the relation between the inner experience and the physiological correlates of inner speech production was only inferred from group means. However, given the previous point, this appears highly problematic. We explore each of these limitations and suggests ways forward in the following section.

@@

```{r general, fig.cap = "Average log-EMG amplitude by muscle and condition. The black dots and intervals represent the by-group average and 95\\% confidence interval (N = 26). The horizontal white line in the violin plot represents the median. The grey dots represent the individual-level average natural logarithm of the EMG amplitude by muscle and condition."}
# position for groups
pd <- position_dodge(0.8)

@@

```{r model, results = "hide"}
# priors for the varying-effects model
priors_varying <- c(
    prior(normal(0, 10), class = Intercept, resp = "FRO"),
    prior(normal(0, 1), class = b, resp = "FRO"),
    prior(exponential(1), class = sigma, resp = "FRO"),
    prior(exponential(1), class = sd, resp = "FRO"),
    prior(normal(0, 10), class = Intercept, resp = "OOS"),
    prior(normal(0, 1), class = b, resp = "OOS"),
    prior(exponential(1), class = sigma, resp = "OOS"),
    prior(exponential(1), class = sd, resp = "OOS"),
    prior(normal(0, 10), class = Intercept, resp = "OOI"),
    prior(normal(0, 1), class = b, resp = "OOI"),
    prior(exponential(1), class = sigma, resp = "OOI"),
    prior(exponential(1), class = sd, resp = "OOI")
    )

@@

## Concluding on the null from low-powered studies is uninformative

@@

## Manipulating rumination within-subject

@@

## Does everyone?

@@

## Relation between self-report and EMG correlates

@@

>>>>

---
title: |
  | Re-analysing the data from @moffatt_inner_2020: A textbook illustration of the absence of evidence fallacy
  | 
  | Submitted to Meta-Psychology. Click here to follow the fully transparent editorial process of this submission. Participate in open peer review by commenting through hypothes.is directly on this preprint.

shorttitle: "The absence of evidence fallacy"

author: 
  - name: Ladislas Nalborczyk
    affiliation: 1
    corresponding: yes
    address: GIPSA-lab, CNRS, Univ. Grenoble Alpes, 11 Rue des Mathématiques, 38400 Saint-Martin-d'Hères, France
    email: ladislas.nalborczyk@gipsa-lab.fr

affiliation:
  - id: 1
    institution: Univ. Grenoble Alpes, CNRS, Grenoble INP, GIPSA-lab, 38000 Grenoble, France

abstract: "@moffatt_inner_2020 reported the results of an experiment (N = 26 in the final sample) comparing the facial (surface) electromyographic correlates of mental rumination and distraction, following an experimentally induced stressor. Based on the absence of significant difference in the perioral muscular activity between the rumination and distraction conditions, @moffatt_inner_2020 concluded that *self-reported* inner experience was unrelated to peripheral muscular activity as assessed using surface electromyography. We suggest this conclusion is hasty and based on waggly evidence. Indeed, concluding on the absence of an effect based on a low-powered non-significant p-value is strongly problematic/uninformative. Moreover, the relation between self-reports and physiological measures was not *directly* assessed, but only indirectly inferred from differences (or absence thereof) in group means. Given the ample inter-individual variability in these measures (as suggested by our reanalysis), we think inferring the individual-level relation between self-reports and physiological measures from group means is inappropriate. Given these limitations, we conclude that it is unclear whether the target article adds to the current/extent knowledge and we suggest ways forward, both from a theoretical and from a methodological perspective. Complete source code, reproducible analyses, and figures are available at https://github.com/lnalborczyk/inner_experience_EMG."

keywords: NHST, Bayesian, fallacy, reanalysis, inner speech, rumination, electromyography
csl: csl/apa7.csl
classoption: man, donotrepeattitle
linenumbers: yes
floatsintext: yes
link-citations: true
numbersections: false

bibliography: [bib/library.bib, bib/r-references.bib]

nocite: | 
  @R-base, @R-knitr, @R-rmarkdown, @R-tidyverse, @R-papaja, @R-wordcountaddin,
  @R-here

output:
  papaja::apa6_pdf:
    keep_tex: true
    highlight: tango
    latex_engine: xelatex

editor_options: 
  chunk_output_type: console
---

```{r setup, include = FALSE, message = FALSE, warning = FALSE}
library(wordcountaddin)
library(BayesFactor)
library(ggbeeswarm)
library(rmarkdown)
library(tidyverse)
library(patchwork)
library(parallel)
library(papaja)
library(readxl)
library(knitr)
library(brms)
library(glue)
library(here)
library(pwr)

@@

NB: You can add comments using these tags

@@

@@

The activity of silently talking to oneself or "inner speech" is a foundational ability, allowing oneself to remember, plan self-motivate or self-regulate. Despite its multiple adaptive functions in everyday life, inner speech can go awry and leads to sustained negative... These inner speech "dysfunctions" [for reviews, see @alderson-day_inner_2015; @perrone-bertolotti_what_2014; @loevenbruck_cognitive_2018]...

Given the predominantly verbal nature of rumination [e.g., @ehring_repetitive_2008; @goldwin_concreteness_2012; @goldwin_concreteness_2013; @mclaughlin_effects_2007], we previously proposed to consider rumination as a form of inner speech and to study it using the methods that have been used historically to study other forms of inner speech, namely, by using surface electromyography and motor interference protocols [e.g., @nalborczyk_orofacial_2017; @nalborczyk_understanding_2019; @nalborczyk_dissociating_2020; @nalborczyk_articulatory_2020]. We first showed that induced rumination was accompanied by increased facial (both over a forehead and a perioral site) muscular activity in comparison to a rest period [@nalborczyk_orofacial_2017]. However, because the rumination condition did not have a proper control condition in this first study, it was unclear whether this perioral activity were specifically related to (inner) speech processes. Therefore, we ran an extension of this study, in which we compared verbal to non-verbal rumination, which suggested that the facial EMG correlates we have previously identified were not specifically related to the verbal content of the ruminative thoughts [@nalborczyk_dissociating_2020]. We discussed these findings in length and proposed several theoretical interpretations that can account for these results in the discussion section of @nalborczyk_dissociating_2020 and more extensively in @nalborczyk_understanding_2019. Although these discussions were blatantly ignored by @moffatt_inner_2020, their experimental design nevertheless had the potential to inform our understanding of the involvement of the speech motor system in different varieties of inner speech as well as to clarify the relation between the peripheral correlates of inner speech and the (self-reported) subjective experience.

The main conclusion from @moffatt_inner_2020 is that inner experience between induced rumination and distraction differs "without a change in electromyographic correlates of inner speech". In other words, they suggest that the subjective experience of inner speech is unrelated (or loosely related) to the electromyographic correlates of inner speech, which are thought to be represented mostly by the EMG amplitude recorded over the OOI and OOS muscles. However, for this in-sample observation to be of interest in an out-of-sample context (i.e., to be informative of other non-observed individuals, or said otherwise, to bring information about the population), this absence of difference has to be based on sufficiently powered sample size (given the target effect size) as well as on reliable measures. This is unlikely to be the case here, for reasons that we will present and discuss in the following. Moreover, a simple visual exploration of the data reveals important variability between individuals in the main effect of interest. That is, some participants had higher perioral (OOS and OOI) muscular activity in the rumination condition than in the distraction condition, and some other participants showed the reverse pattern. This suggests unexplored variation in the determinants of this effect (e.g., the content of the inner experience). Indeed, the relation between the inner experience and the physiological correlates of inner speech production was only inferred from group means. However, given the previous point, this appears highly problematic. We explore each of these limitations and suggests ways forward in the following section.

@@

@moffatt_inner_2020 recorded... in 26 participants (data available at https://osf.io/hj7tz/)... The EMG data is depicted in Figure \@ref(fig:general) by condition (where `BAS`, `DIS`, and `RUM` refer to the baseline, distraction, and rumination conditions, respectively) and by muscle (`FRO`, `OOI`, `OOS`).

```{r general, fig.cap = "Average natural logarithm of the EMG peak amplitude per muscle and condition. The black dots and intervals represent the by-group average and 95\\% confidence interval (N = 26). The horizontal white line in the violin plot represents the median. The grey dots represent the individual-level average natural logarithm of the EMG amplitude by muscle and condition."}
# position for groups
pd <- position_dodge(0.8)

@@

We fitted a multivariate Bayesian regression model with varying-intercepts (by participant) and weakly informative priors on these data using the `brms` package [@R-brms]...

```{r models, results = "hide"}
##############################
# Bayesian multilevel models #
##############################

# priors for the constant-effects model
priors_constant <- c(
    prior(normal(0, 10), class = Intercept, resp = "FRO"),
    prior(normal(0, 1), class = b, resp = "FRO"),
    prior(exponential(1), class = sigma, resp = "FRO"),
    prior(normal(0, 10), class = Intercept, resp = "OOS"),
    prior(normal(0, 1), class = b, resp = "OOS"),
    prior(exponential(1), class = sigma, resp = "OOS"),
    prior(normal(0, 10), class = Intercept, resp = "OOI"),
    prior(normal(0, 1), class = b, resp = "OOI"),
    prior(exponential(1), class = sigma, resp = "OOI")
    )

# fitting the model
constant_effects <- brm(
    mvbind(FRO, OOS, OOI) ~ 1 + condition,
    family = gaussian(),
    prior = priors_constant,
    data = df2,
    chains = 4, cores = detectCores(),
    warmup = 2000, iter = 10000,
    control = list(adapt_delta = 0.95),
    sample_prior = TRUE,
    file = here("models/constant_effects")
    )

# priors for the varying-effects model
priors_varying <- c(
    prior(normal(0, 10), class = Intercept, resp = "FRO"),
    prior(normal(0, 1), class = b, resp = "FRO"),
    prior(exponential(1), class = sigma, resp = "FRO"),
    prior(exponential(1), class = sd, resp = "FRO"),
    prior(normal(0, 10), class = Intercept, resp = "OOS"),
    prior(normal(0, 1), class = b, resp = "OOS"),
    prior(exponential(1), class = sigma, resp = "OOS"),
    prior(exponential(1), class = sd, resp = "OOS"),
    prior(normal(0, 10), class = Intercept, resp = "OOI"),
    prior(normal(0, 1), class = b, resp = "OOI"),
    prior(exponential(1), class = sigma, resp = "OOI"),
    prior(exponential(1), class = sd, resp = "OOI")
    )

@@

```{r posterior, eval = FALSE, fig.cap = "..."}
# extracts posterior samples
post <- posterior_samples(varying_effects)

plotPost(
    post$b_FRO_conditionRUM - post$b_FRO_conditionDIS,
    compVal = 0#, ROPE = c(-0.1, 0.1)
    )

plotPost(
    post$b_OOI_conditionRUM - post$b_OOI_conditionDIS,
    compVal = 0#, ROPE = c(-0.1, 0.1)
    )

plotPost(
    post$b_OOS_conditionRUM - post$b_OOS_conditionDIS,
    compVal = 0#, ROPE = c(-0.1, 0.1)
    )
```

```{r bf, fig.width = 10, fig.cap = "Savage-Dickey Bayes factor for the difference between the rumination and distraction conditions for each muscle. The BF is computed as the ratio of the posterior density to the prior density at $\\theta = 0$."}
# BF testing
hyp_fro <- hypothesis(
    varying_effects,
    "FRO_conditionDIS = FRO_conditionRUM"
    )

p1 <- plot(hyp_fro, plot = FALSE, theme = theme_bw(base_size = 12) )[[1]] +
    geom_vline(xintercept = 0, linetype = 2) +
    coord_cartesian(xlim = c(-5, 5) ) +
    theme_bw(base_size = 10) +
    theme(legend.title = element_blank() ) +
    labs(
      x = "Log EMG peak amplitude",
      title = paste0(
        "BF10 = ", round(1 / hyp_fro$hypothesis$Evid.Ratio, 3),
        " (BF01 = ", round(hyp_fro$hypothesis$Evid.Ratio, 3), ")"
        )
      )

hyp_ooi <- hypothesis(
    varying_effects,
    "OOI_conditionDIS = OOI_conditionRUM"
    )

p2 <- plot(hyp_ooi, plot = FALSE, theme = theme_bw(base_size = 12) )[[1]] +
    geom_vline(xintercept = 0, linetype = 2) +
    coord_cartesian(xlim = c(-5, 5) ) +
    theme_bw(base_size = 10) +
    theme(legend.title = element_blank() ) +
    labs(
      x = "Log EMG peak amplitude",
      title = paste0(
        "BF10 = ", round(1 / hyp_ooi$hypothesis$Evid.Ratio, 3),
        " (BF01 = ", round(hyp_ooi$hypothesis$Evid.Ratio, 3), ")"
        )
      )

hyp_oos <- hypothesis(
    varying_effects,
    "OOS_conditionDIS = OOS_conditionRUM"
    )

p3 <- plot(hyp_oos, plot = FALSE, theme = theme_bw(base_size = 12) )[[1]] +
    geom_vline(xintercept = 0, linetype = 2) +
    coord_cartesian(xlim = c(-5, 5) ) +
    theme_bw(base_size = 10) +
    theme(legend.title = element_blank() ) +
    labs(
      x = "Log EMG peak amplitude",
      title = paste0(
        "BF10 = ", round(1 / hyp_oos$hypothesis$Evid.Ratio, 3),
        " (BF01 = ", round(hyp_oos$hypothesis$Evid.Ratio, 3), ")"
        )
      )

p1 + p2 + p3 + plot_layout(guides = "collect") & theme(legend.position = "bottom")
```

A summary of the estimations from this model is presented in Table \@ref(tab:summary). For each muscle, the intercept gives the estimated of the natural logarithm of the EMG peak amplitude in the baseline condition, whereas the `conditionDIS` and `conditionRUM` parameters indicate deviations from the baseline in the distraction and rumination conditions (respectively).

```{r summary, result = "asis"}
summary_varying_effects <- 
    # retrieve estimations from bmod5
    summary(varying_effects, prob = 0.95)$fixed %>%
    # convert to dataframe
    data.frame %>%
    # sets rownames as the first column
    rownames_to_column(var = "response") %>%
    # removing ESS information
    dplyr::select(-Bulk_ESS, -Tail_ESS) %>%
    # split response name as two columns
    # separate(col = response, into = c("response", "term"), sep = "_", extra = "merge") %>%
    # rename columns
    magrittr::set_colnames(
        c("Term", "Estimate", "SE", "Lower", "Upper", "Rhat")
        ) %>%
    # compute BF for each effect
    mutate(
        BF10 = 1 / hypothesis(varying_effects, glue("{Term} = 0") )$hypothesis$Evid.Ratio
        ) %>%
    # round BFs
    # mutate(BF01 = ifelse(BF01 < .001, "<0.001", BF01) ) %>%
    mutate(BF10 = ifelse(BF10 < .001, changeSciNot(BF10), round(BF10, 3) ) ) %>%
    # if numeric, round to 3 decimals
    mutate_if(is.numeric, round, 3)

apa_table(
    summary_varying_effects,
    placement = "H",
    align = c("l", rep("c", 6) ),
    caption = "Estimated value of the natural logarithm of the EMG peak amplitude in each condition and for each muscle.",
    note = "For each effect, the 'Estimate' reports the estimated average value of the natural logarithm of the EMG peak amplitude, followed by its standard error (SE). The 'Lower' and 'Upper' columns contain the lower and upper bounds of the 95% CrI, whereas the 'Rhat' column reports the Gelman-Rubin statistic. The last column reports the BF in favour of the alternative hypothesis (relative to the null hypothesis).",
    small = TRUE,
    escape = TRUE
    )
```

## Concluding on the null from low-powered studies: what could go wrong?

@@

## Within-subject manipulation of rumination and distraction

@@

## Does everyone show the effect?

@@

## Relating the subjective inner experience to the psychophysiological correlates

@@

Acknowledgements will be included in the final version of this manuscript.

### 2020-09-30 manuscript.tex
minor updates in manuscript

  man,floatsintext]{apa6}

@@

  pdftitle={Re-analysing the data from Moffatt et al. (2020): A textbook illustration of the absence of evidence fallacy},

@@

\title{Re-analysing the data from Moffatt et al. (\protect\hyperlink{ref-moffatt_inner_2020}{2020}): A textbook illustration of the absence of evidence fallacy}

@@

Wordcount (excluding abstract, references, tables, and figures): 1727

@@

The activity of silently talking to oneself or ``inner speech'' is a foundational ability\ldots{} despite its multiple adaptive functions in everyday life, inner speech can go awry and leads to sustained negative\ldots{} These inner speech ``dysfunctions'' (for reviews, see Alderson-Day \& Fernyhough, \protect\hyperlink{ref-alderson-day_inner_2015}{2015}; Lœvenbruck et al., \protect\hyperlink{ref-loevenbruck_cognitive_2018}{2018}; Perrone-Bertolotti et al., \protect\hyperlink{ref-perrone-bertolotti_what_2014}{2014})\ldots{}

Given the predominantly verbal nature of rumination {[}{]}, we previously proposed to study rumination as other forms of inner speech have been studied in the past, namely using surface electromyography and motor interference protocols (e.g., Nalborczyk et al., \protect\hyperlink{ref-nalborczyk_orofacial_2017}{2017}; Nalborczyk, \protect\hyperlink{ref-nalborczyk_understanding_2019}{2019}; Nalborczyk, Perrone-Bertolotti, et al., \protect\hyperlink{ref-nalborczyk_articulatory_2020}{2020}; Nalborczyk, Grandchamp, et al., \protect\hyperlink{ref-nalborczyk_can_2020}{2020})\ldots{}

We have previously shown that\ldots{} However, it was unclear\ldots{} therefore, the extension of our study by Moffatt et al. (\protect\hyperlink{ref-moffatt_inner_2020}{2020}), which consisted by including a distraction control group, is more than welcome\ldots{} However\ldots{}

The main conclusion from Moffatt et al. (\protect\hyperlink{ref-moffatt_inner_2020}{2020}) is that inner experience between induced rumination and distraction differs ``without a change in electromyographic correlates of inner speech''. In other words, their conclusion is that inner experience is unrelated (or loosely related) to the electromyographic correlates of inner speech, which are thought to be represented mostly by the EMG amplitude recorded over the OOI and OOS muscles. However, for this in-sample observation to be of interest in an out-of-sample context (i.e., to be informative of other non-observed individuals, or said otherwise, to bring information about the population), this absence of difference has to be based on sufficiently powered sample size (given the target effect size) as well as on reliable measures. Moreover, a simple visual exploration of the data reveals important variability between individuals in the main effect of interest. That is, some participants had higher perioral (OOS and OOI) muscular activity in the rumination condition than in the distraction condition, and some other participants showed the reverse pattern. This suggests unexplored variation in the determinants of this effects (e.g., the content of the inner experience). Indeed, the relation between the inner experience and the physiological correlates of inner speech production was only inferred from group means. However, given the previous point, this appears highly problematic. We explore each of these limitations and suggests ways forward in the following section.

@@

@@

\caption{Average log-EMG amplitude by muscle and condition. The black dots and intervals represent the by-group average and 95\% confidence interval (N = 26). The horizontal white line in the violin plot represents the median. The grey dots represent the individual-level average natural logarithm of the EMG amplitude by muscle and condition.}\label{fig:general}
\end{figure}

@@

\hypertarget{concluding-on-the-null-from-low-powered-studies-is-uninformative}{%
\subsection{Concluding on the null from low-powered studies is uninformative}\label{concluding-on-the-null-from-low-powered-studies-is-uninformative}}

@@

\hypertarget{manipulating-rumination-within-subject}{%
\subsection{Manipulating rumination within-subject}\label{manipulating-rumination-within-subject}}

@@

\hypertarget{does-everyone}{%
\subsection{Does everyone?}\label{does-everyone}}

@@

\hypertarget{relation-between-self-report-and-emg-correlates}{%
\subsection{Relation between self-report and EMG correlates}\label{relation-between-self-report-and-emg-correlates}}

@@

@@

>>>>

  man, donotrepeattitle,floatsintext]{apa6}

@@

@@

\title{Re-analysing the data from Moffatt et al. (\protect\hyperlink{ref-moffatt_inner_2020}{2020}): A textbook illustration of the absence of evidence fallacy\\
~\\
Submitted to Meta-Psychology. Click here to follow the fully transparent editorial process of this submission. Participate in open peer review by commenting through hypothes.is directly on this preprint.}

@@

@@

The activity of silently talking to oneself or ``inner speech'' is a foundational ability, allowing oneself to remember, plan self-motivate or self-regulate. Despite its multiple adaptive functions in everyday life, inner speech can go awry and leads to sustained negative\ldots{} These inner speech ``dysfunctions'' (for reviews, see Alderson-Day \& Fernyhough, \protect\hyperlink{ref-alderson-day_inner_2015}{2015}; Lœvenbruck et al., \protect\hyperlink{ref-loevenbruck_cognitive_2018}{2018}; Perrone-Bertolotti et al., \protect\hyperlink{ref-perrone-bertolotti_what_2014}{2014})\ldots{}

Given the predominantly verbal nature of rumination (e.g., Ehring \& Watkins, \protect\hyperlink{ref-ehring_repetitive_2008}{2008}; Goldwin et al., \protect\hyperlink{ref-goldwin_concreteness_2013}{2013}; Goldwin \& Behar, \protect\hyperlink{ref-goldwin_concreteness_2012}{2012}; McLaughlin et al., \protect\hyperlink{ref-mclaughlin_effects_2007}{2007}), we previously proposed to consider rumination as a form of inner speech and to study it using the methods that have been used historically to study other forms of inner speech, namely, by using surface electromyography and motor interference protocols (e.g., Nalborczyk et al., \protect\hyperlink{ref-nalborczyk_orofacial_2017}{2017}; Nalborczyk, \protect\hyperlink{ref-nalborczyk_understanding_2019}{2019}; Nalborczyk, Perrone-Bertolotti, et al., \protect\hyperlink{ref-nalborczyk_articulatory_2020}{2020}; Nalborczyk, Banjac, et al., \protect\hyperlink{ref-nalborczyk_dissociating_2020}{2020}). We first showed that induced rumination was accompanied by increased facial (both over a forehead and a perioral site) muscular activity in comparison to a rest period (Nalborczyk et al., \protect\hyperlink{ref-nalborczyk_orofacial_2017}{2017}). However, because the rumination condition did not have a proper control condition in this first study, it was unclear whether this perioral activity were specifically related to (inner) speech processes. Therefore, we ran an extension of this study, in which we compared verbal to non-verbal rumination, which suggested that the facial EMG correlates we have previously identified were not specifically related to the verbal content of the ruminative thoughts (Nalborczyk, Banjac, et al., \protect\hyperlink{ref-nalborczyk_dissociating_2020}{2020}). We discussed these findings in length and proposed several theoretical interpretations that can account for these results in the discussion section of Nalborczyk, Banjac, et al. (\protect\hyperlink{ref-nalborczyk_dissociating_2020}{2020}) and more extensively in Nalborczyk (\protect\hyperlink{ref-nalborczyk_understanding_2019}{2019}). Although these discussions were blatantly ignored by Moffatt et al. (\protect\hyperlink{ref-moffatt_inner_2020}{2020}), their experimental design nevertheless had the potential to inform our understanding of the involvement of the speech motor system in different varieties of inner speech as well as to clarify the relation between the peripheral correlates of inner speech and the (self-reported) subjective experience.

The main conclusion from Moffatt et al. (\protect\hyperlink{ref-moffatt_inner_2020}{2020}) is that inner experience between induced rumination and distraction differs ``without a change in electromyographic correlates of inner speech''. In other words, they suggest that the subjective experience of inner speech is unrelated (or loosely related) to the electromyographic correlates of inner speech, which are thought to be represented mostly by the EMG amplitude recorded over the OOI and OOS muscles. However, for this in-sample observation to be of interest in an out-of-sample context (i.e., to be informative of other non-observed individuals, or said otherwise, to bring information about the population), this absence of difference has to be based on sufficiently powered sample size (given the target effect size) as well as on reliable measures. This is unlikely to be the case here, for reasons that we will present and discuss in the following. Moreover, a simple visual exploration of the data reveals important variability between individuals in the main effect of interest. That is, some participants had higher perioral (OOS and OOI) muscular activity in the rumination condition than in the distraction condition, and some other participants showed the reverse pattern. This suggests unexplored variation in the determinants of this effect (e.g., the content of the inner experience). Indeed, the relation between the inner experience and the physiological correlates of inner speech production was only inferred from group means. However, given the previous point, this appears highly problematic. We explore each of these limitations and suggests ways forward in the following section.

@@

Moffatt et al. (\protect\hyperlink{ref-moffatt_inner_2020}{2020}) recorded\ldots{} in 26 participants (data available at \url{https://osf.io/hj7tz/})\ldots{} The EMG data is depicted in Figure \ref{fig:general} by condition (where \texttt{BAS}, \texttt{DIS}, and \texttt{RUM} refer to the baseline, distraction, and rumination conditions, respectively) and by muscle (\texttt{FRO}, \texttt{OOI}, \texttt{OOS}).

@@

\caption{Average natural logarithm of the EMG peak amplitude per muscle and condition. The black dots and intervals represent the by-group average and 95\% confidence interval (N = 26). The horizontal white line in the violin plot represents the median. The grey dots represent the individual-level average natural logarithm of the EMG amplitude by muscle and condition.}\label{fig:general}
\end{figure}

We fitted a multivariate Bayesian regression model with varying-intercepts (by participant) and weakly informative priors on these data using the \texttt{brms} package (Bürkner, \protect\hyperlink{ref-R-brms}{2017})\ldots{}

{\centering \includegraphics[width=1\linewidth]{manuscript_files/figure-latex/bf-1}

\caption{Savage-Dickey Bayes factor for the difference between the rumination and distraction conditions for each muscle. The BF is computed as the ratio of the posterior density to the prior density at $\theta = 0$.}\label{fig:bf}
\end{figure}

A summary of the estimations from this model is presented in Table \ref{tab:summary}. For each muscle, the intercept gives the estimated of the natural logarithm of the EMG peak amplitude in the baseline condition, whereas the \texttt{conditionDIS} and \texttt{conditionRUM} parameters indicate deviations from the baseline in the distraction and rumination conditions (respectively).

\begin{table}[H]

\begin{center}
\begin{threeparttable}

\caption{\label{tab:summary}Estimated value of the natural logarithm of the EMG peak amplitude in each condition and for each muscle.}

\small{

\begin{tabular}{lcccccc}
\toprule
Term & \multicolumn{1}{c}{Estimate} & \multicolumn{1}{c}{SE} & \multicolumn{1}{c}{Lower} & \multicolumn{1}{c}{Upper} & \multicolumn{1}{c}{Rhat} & \multicolumn{1}{c}{BF10}\\
\midrule
FRO\_Intercept & -2.08 & 0.10 & -2.27 & -1.89 & 1.00 & NA\\
OOS\_Intercept & -2.36 & 0.14 & -2.64 & -2.09 & 1.00 & NA\\
OOI\_Intercept & -2.29 & 0.12 & -2.52 & -2.05 & 1.00 & NA\\
FRO\_conditionDIS & 0.00 & 0.07 & -0.14 & 0.13 & 1.00 & 0.06\\
FRO\_conditionRUM & 0.22 & 0.07 & 0.09 & 0.36 & 1.00 & 10.51\\
OOS\_conditionDIS & 0.16 & 0.11 & -0.06 & 0.38 & 1.00 & 0.35\\
OOS\_conditionRUM & 0.21 & 0.11 & -0.01 & 0.43 & 1.00 & 0.71\\
OOI\_conditionDIS & 0.29 & 0.12 & 0.06 & 0.53 & 1.00 & 2.31\\
OOI\_conditionRUM & 0.37 & 0.12 & 0.14 & 0.61 & 1.00 & 15.22\\
\bottomrule
\addlinespace
\end{tabular}

\begin{tablenotes}[para]
\normalsize{\textit{Note.} For each effect, the 'Estimate' reports the estimated average value of the natural logarithm of the EMG peak amplitude, followed by its standard error (SE). The 'Lower' and 'Upper' columns contain the lower and upper bounds of the 95\% CrI, whereas the 'Rhat' column reports the Gelman-Rubin statistic. The last column reports the BF in favour of the alternative hypothesis (relative to the null hypothesis).}
\end{tablenotes}

\end{threeparttable}
\end{center}

\end{table}

@@

\hypertarget{concluding-on-the-null-from-low-powered-studies-what-could-go-wrong}{%
\subsection{Concluding on the null from low-powered studies: what could go wrong?}\label{concluding-on-the-null-from-low-powered-studies-what-could-go-wrong}}

@@

\hypertarget{within-subject-manipulation-of-rumination-and-distraction}{%
\subsection{Within-subject manipulation of rumination and distraction}\label{within-subject-manipulation-of-rumination-and-distraction}}

@@

\hypertarget{does-everyone-show-the-effect}{%
\subsection{Does everyone show the effect?}\label{does-everyone-show-the-effect}}

@@

\hypertarget{relating-the-subjective-inner-experience-to-the-psychophysiological-correlates}{%
\subsection{Relating the subjective inner experience to the psychophysiological correlates}\label{relating-the-subjective-inner-experience-to-the-psychophysiological-correlates}}

@@

\leavevmode\hypertarget{ref-ehring_repetitive_2008}{}%
Ehring, T., \& Watkins, E. R. (2008). Repetitive negative thinking as a transdiagnostic process. \emph{International Journal of Cognitive Therapy}, \emph{1}(3), 192--205. \url{https://doi.org/10.1680/ijct.2008.1.3.192}

\leavevmode\hypertarget{ref-goldwin_concreteness_2012}{}%
Goldwin, M., \& Behar, E. (2012). Concreteness of idiographic periods of worry and depressive rumination. \emph{Cognitive Therapy and Research}, \emph{36}(6), 840--846. \url{https://doi.org/10.1007/s10608-011-9428-1}

\leavevmode\hypertarget{ref-goldwin_concreteness_2013}{}%
Goldwin, M., Behar, E., \& Sibrava, N. J. (2013). Concreteness of depressive rumination and trauma recall in individuals with elevated trait rumination and/or posttraumatic stress symptoms. \emph{Cognitive Therapy and Research}, \emph{37}(4), 680--689. \url{https://doi.org/10.1007/s10608-012-9507-y}

@@

\leavevmode\hypertarget{ref-mclaughlin_effects_2007}{}%
McLaughlin, K. A., Borkovec, T. D., \& Sibrava, N. J. (2007). The effects of worry and rumination on affect states and cognitive activity. \emph{Behavior Therapy}, \emph{38}(1), 23--38. \url{https://doi.org/10.1016/j.beth.2006.03.003}
